Automatic prostate cancer detection through \acs*{dce}-\acs*{mri} images: all you need is a good normalization

### 2017-04-29 01_introduction/introduction.tex
Move to IEEE template

\Ac{cap} is the second most frequently diagnosed cancer in men,
accounting for 899,000 cases and leading to 258,100 deaths per year~\citep{ferlay2010estimates}.
As highlighted by the PI-RADS Steering Committee, the two main challenges to be addressed are~\citep{weinreb2016pi}:
(i) improving the detection of clinically significant \ac{cap} and
(ii) increasing confidence in benign or dormant cases and therefore avoiding unnecessary invasive medical exams.
In this regard, \ac{mpmri} is frequently used to build robust \ac{cad} systems to detect, localize, and grade \ac{cap}.
In general, \ac{cad} systems are based on \ac{mpmri} which potentially combines several of the following modalities~\citep{lemaitre2015computer}: \ac{t2w}-\ac{mri}, \ac{dce}-\ac{mri}, \ac{adc} maps, and \ac{mrsi}.

In \ac{dce}-\ac{mri}, a contrast media is injected intravenously and a set of images is acquired over time.
Consequently, each voxel in an image corresponds to a dynamic signal that is related to both contrast agent concentration and the vascular properties of the tissue.
Therefore, changes in the enhanced signal allows for the
discrimination of healthy tissues from \ac{cap} tissues.
In fact, these properties are automatically extracted using quantitative or semi-quantitative approaches~\citep{lemaitre2015computer}.

\emph{Quantitative} approaches uses pharmacokinetic modelling based on a bicompartment model, namely Brix~\citep{brix1991pharmacokinetic} and Tofts~\citep{tofts1995quantitative} models.
The parameters of the Brix model are inferred by assuming a linear relationship between the media concentration and the \ac{mri} signal intensity.
However, this assumption has been shown to lead to inaccurate
estimations of the pharmacokinetic parameters~\citep{heilmann2006determination}.
In contrast, the Tofts model requires the conversion of \ac{mri}
signal intensity to concentration, which becomes a non-linear
relationship using a specific equation of \ac{mri} sequences (e.g., FLASH sequence).
Tofts modeling, however, is highly complex~\citep{gliozzi2011phenomenological}.
Achieving the conversion using the non-linear approach requires the
acquisition of a T$_1$ map which is not always possible during clinical examination.
Additionally, the parameter calculation requires the \ac{aif} which is challenging to measure and can also lead to an inaccurate estimation.

\emph{Semi-quantitative} approaches are mathematical rather than pharmacokinetic because no pharmacokinetic assumptions regarding the relationship between the \ac{mri} signal and the contrast agent are made~\citep{huisman2001accurate,gliozzi2011phenomenological}.
These methods are advantageous because they do not require any knowledge of the \ac{mri} sequence or any conversion from signal intensity to concentration.
However, they present some limitations: the heuristic approach
proposed by \citeauthor{huisman2001accurate} requires an initial
estimate of the standard deviation of the signal noise and some manual tuning.

>>>>

\Ac{cap} is the second most frequently diagnosed cancer in men,
accounting for 899,000 cases and leading to 258,100 deaths per year~\cite{ferlay2010estimates}.
As highlighted by the PI-RADS Steering Committee, the two main challenges to be addressed are~\cite{weinreb2016pi}:
(i) improving the detection of clinically significant \ac{cap} and
(ii) increasing confidence in benign or dormant cases and therefore avoiding unnecessary invasive medical exams.
In this regard, \ac{mpmri} is frequently used to build robust \ac{cad} systems to detect, localize, and grade \ac{cap}.
In general, \ac{cad} systems are based on \ac{mpmri} which potentially combines several of the following modalities~\cite{lemaitre2015computer}: \ac{t2w}-\ac{mri}, \ac{dce}-\ac{mri}, \ac{adc} maps, and \ac{mrsi}.

In \ac{dce}-\ac{mri}, a contrast media is injected intravenously and a set of images is acquired over time.
Consequently, each voxel in an image corresponds to a dynamic signal that is related to both contrast agent concentration and the vascular properties of the tissue.
Therefore, changes in the enhanced signal allows for the
discrimination of healthy tissues from \ac{cap} tissues.
In fact, these properties are automatically extracted using quantitative or semi-quantitative approaches~\cite{lemaitre2015computer}.

\emph{Quantitative} approaches uses pharmacokinetic modelling based on a bicompartment model, namely Brix~\cite{brix1991pharmacokinetic} and Tofts~\cite{tofts1995quantitative} models.
The parameters of the Brix model are inferred by assuming a linear relationship between the media concentration and the \ac{mri} signal intensity.
However, this assumption has been shown to lead to inaccurate
estimations of the pharmacokinetic parameters~\cite{heilmann2006determination}.
In contrast, the Tofts model requires the conversion of \ac{mri}
signal intensity to concentration, which becomes a non-linear
relationship using a specific equation of \ac{mri} sequences (e.g., FLASH sequence).
Tofts modeling, however, is highly complex~\cite{gliozzi2011phenomenological}.
Achieving the conversion using the non-linear approach requires the
acquisition of a T$_1$ map which is not always possible during clinical examination.
Additionally, the parameter calculation requires the \ac{aif} which is challenging to measure and can also lead to an inaccurate estimation.

\emph{Semi-quantitative} approaches are mathematical rather than pharmacokinetic because no pharmacokinetic assumptions regarding the relationship between the \ac{mri} signal and the contrast agent are made~\cite{huisman2001accurate,gliozzi2011phenomenological}.
These methods are advantageous because they do not require any knowledge of the \ac{mri} sequence or any conversion from signal intensity to concentration.
However, they present some limitations: the heuristic approach
proposed by \citeauthor{huisman2001accurate} requires an initial
estimate of the standard deviation of the signal noise and some manual tuning \cite{huisman2001accurate}.

### 2017-04-29 02_methods/methods.tex
Update everything
Move to IEEE template

In this work, we propose a method to normalize \ac{dce}-\ac{mri}
prostate data to reduce inter-patient variations, although this method can be applied to any \ac{dce}-\ac{mri} sequences.
In \ac{t2w}-\ac{mri}, these variations are characterized by a shift and a scaling of the intensities as illustrated by the intensity \ac{pdf} in Fig.\,\ref{fig:t2w}.
Therefore, these variations can be corrected using a $z$-score approach,--- i.e., normalizing the data by subtracting the mean and dividing by the standard deviation ---assuming that the data follow a specific distribution~\citep{lemaitre2016normalization}.

\begin{figure*}
  \centering
  \hspace*{\fill}
  \subfigure[]{\label{subfig:pathhist}\includegraphics[width=1\textwidth]{02_methods/figures/heatmaprep.pdf}} \hfill
  \hspace*{\fill}
  \\
  \hspace*{\fill}
  \subfigure[$I_0$: 117; $t_0$: 6\textsuperscript{th} serie; wider st. dev.]{\label{subfig:pat1}\includegraphics[width=.49\textwidth]{02_methods/figures/pat1_annotated.pdf}} \hfill
  \subfigure[$I_0$: 103; $t_0$: 4\textsuperscript{th} serie; narrower std. dev.]{\label{subfig:pat2}\includegraphics[width=.49\textwidth]{02_methods/figures/pat2_annotated.pdf}} \hfill
  \hspace*{\fill}
  \caption{\subref{subfig:pathhist} Illustration of the heatmap
    representation: all \ac*{pdf}s of the prostate gland are
    concatenated together to build an heatmap;
    \subref{subfig:pat1}-\subref{subfig:pat2} Heatmap of 2 patients
    revealing  the three types of inter-patient variations: intensity shift
    ($\Delta_i$), time shift ($\Delta_t$), and intensity scale
    ($\alpha_i$).}
  \label{fig:heatmap}
\end{figure*}

@@

\begin{equation}
  w_{ij} = \begin{cases}
    \alpha \exp(1 - \frac{H(i)}{\max(H)})       & \text{if } x_j = x_i + 1 \text{ and } y_j = y_i, \\
    (1 - \alpha) \exp(1 - \frac{H(i)}{\max(H)}) & \text{if } x_j = x_i \text{ and } y_j = y_i + 1, \\
    0                                           & \text{otherwise},
  \end{cases}
  \label{eq:weight}
\end{equation}

@@

\begin{figure*}
  \centering
  \hspace*{\fill}
  \subfigure[\acs*{rmse} computed for each patient of our dataset.]{\label{fig:rmse}\includegraphics[width=.44\textwidth]{02_methods/figures/rmse.pdf}} \hfill
  \subfigure[\acs*{rmse} after alignment using the curve parametric model.]{\label{fig:rmseal}\includegraphics[width=.49\textwidth]{02_methods/figures/rmse_aligned.pdf}}
  \hspace*{\fill}
  \caption{Illustration of the correction of the time offset and the data dispersion.}
  \label{fig:curveal}
\end{figure*}

@@

In this section, we summarize the different methods that have been
used for the quantification of \ac{dce}-\ac{mri} for \ac{cap}
detection~\citep{lemaitre2015computer} and will be used for comparison in this work.
Furthermore, we would like to emphasize the following additional contributions for this section: (i) a novel automatic \ac{aif} estimation algorithm based on clustering and (ii) a simplified semi-quantitative method using constrained optimization.

@@

In the Brix model~\citep{brix1991pharmacokinetic}, the \ac{mri} signal intensity is assumed to be proportional to the media concentration.
Therefore, the model is expressed as shown in Eq.\,\eqref{eq:brix}, as
follows:

\begin{equation}
  s_n(t) = 1 + A \left[ \frac{\exp(k_{el} t') - 1}{k_{ep}(k_{ep} - k_{el})} \exp(- k_{el} t) - \frac{\exp(k_{ep} t') - 1}{k_{el}(k_{ep} - k_{el})} \exp(- k_{ep} t) \right],
  \label{eq:brix}
\end{equation}

@@

\begin{equation}
  \small
  s_n(t) = 1 + \frac{A}{\tau} \left[ \frac{k_{ep} \left( \exp(k_{el} t') - 1 \right)}{k_{el}(k_{ep} - k_{el})} \exp(- k_{el} t) - \frac{\exp(k_{ep} t') - 1}{(k_{ep} - k_{el})} \exp(- k_{ep} t) \right] ,
  \label{eq:hoffmann}
\end{equation}

@@

\begin{figure*}
  \centering
  \hspace*{\fill}
  \subfigure[Original image.]{\label{fig:org}\includegraphics[width=.3\textwidth]{02_methods/figures/original.pdf}} \hfill
  \subfigure[Candidates region after clustering.]{\label{fig:cand}\includegraphics[width=.3\textwidth]{02_methods/figures/candidate.pdf}} \hfill
  \subfigure[Regions selected after applying the different criteria.]{\label{fig:final}\includegraphics[width=.3\textwidth]{02_methods/figures/aif.pdf}}
  \hspace*{\fill}
  \caption{Illustration of the segmentation of the area used to determine the \acs*{aif}.}
  \label{fig:aif}
\end{figure*}

\begin{description}
  \item[Segmentation of artery voxels and patient-based \ac{aif} estimation] The \ac{aif} signal from \ac{dce}-\ac{mri} can be manually estimated by selecting the most-enhanced voxels from the femoral or iliac arteries~\citep{meng2010comparison}.
    Few methods have been proposed to address the automated extraction
    of the \ac{aif} signal.
    \citeauthor{chen2008automatic} successively filtered the possible
    candidates to be considered as the \ac{aif} such that~\citep{chen2008automatic}:
    (i) dynamic signals with a small peak and voxels with a small wash-in are rejected by thresholding,
    (ii) a blob detector is used and large enough regions are maintained, and
    (iii) circular and cylindricality criteria are used to reject the false positives.
    \citeauthor{zhu2011automated} proposed an iterative method that
    involves the selection of voxels that best fit a gamma variate function~\citep{zhu2011automated}.
    However, this method requires the computation of the first and
    second derivatives as well as the maximum curvature points.
    \citeauthor{shanbhag2012generalized} proposed the following 4-step algorithm~\citep{shanbhag2012generalized,fennessy2015quantitative}:
    (i) remove the slices with artifacts and find the best slices based on intrinsic anatomic landmarks and enhancement characteristics,
    (ii) find the voxel candidates using the maximum enhanced voxels and a multi-label maximum entropy based thresholding algorithm,
    (iii) exclude the region next to the endorectal coil, and
    (iv) select the best 5 candidates that meet the enhancement characteristics and are correlated.

    All the above methods are rather complex, therefore,  we propose a simpler method that is based on the following reasonable assumptions:
    (i) all possible \ac{aif} signal candidates should have a similar shape,
    (ii) they should all have high enhancement, and
    (iii) the arteries should be almost round and within a size range.
    Therefore, each slice is clustered into regions using K-means clustering with $k=6$.
    The cluster made of the most enhanced signals is selected since it contains the artery signals.
    In this regards, the selection criteria corresponds to the 90\textsuperscript{th} percentile of the maximum \ac{dce}-\ac{mri} signal.
    Finally, regions with an eccentricity smaller than $0.5$ and an area in the range of $[100, 400]$ voxels are kept.
    Additionally, to remove voxels contaminated by the partial volume effect, only the $10\%$ most enhanced voxels of the possible candidates are kept as proposed by~\citep{schabel2008uncertainty} and the average signal is computed.
    A summary of the different segmentation steps is presented in Fig.\,\ref{fig:aif}.
    \item[Conversion of \ac{mri} signal intensity to concentration] To estimate the free parameters of the Tofts model (see Eq.\,\eqref{eq:exttofts}), the concentrations $C_t(t)$ and $C_p(t)$ need to be computed from the \ac{mri} signal intensity and the \ac{aif} signal, respectively.
      This conversion is based on the equation of the FLASH sequence\textemdash see~\ref{app:signaltoconc} for details\textemdash and is formulated as in Eq.\,\eqref{eq:conv}:
      \begin{equation}
        c(t) = \frac{1}{TR \cdot r_1} \ln\left( \frac{1 - \cos \alpha \cdot S^{*}\frac{s(t)}{S_0}}{1 - S^{*}\frac{s(t)}{S_0}} \right) - \frac{R_{10}}{r_1} ,
        \label{eq:conv}
      \end{equation}
      \noindent with,
      \begin{equation}
        S^{*} = \frac{1 - \exp(- TR \cdot R_{10})}{1 - \cos \alpha \cdot \exp(- TR \cdot R_{10})} ,
        \label{eq:sstarconv}
      \end{equation}
      \noindent where $s(t)$ is the \ac{mri} signal, $S_0$ is the \ac{mri} signal prior to the injection of the contrast media, $\alpha$ is the flip angle, $TR$ is the \acf{tr}, $R_{10}$ is the pre-contrast tissue relaxation time also equal to $\frac{1}{T_{10}}$, and $r_1$ is the relaxivity coefficient of the contrast agent.

      $T_{10}$ can be estimated from the acquisition of a T$_1$ map.
      However, this modality is not part of the clinical trial in this research and the value of $T_{10}$ is fixed to \SI{1600}{\ms} for both blood and prostate, in accordance with the values found in the literature~\citep{fennessy2015quantitative,de2004mr,carr2011magnetic}.
      \item[Estimation of population-based \ac{aif}] While estimating
        the pharmacokinetic parameters using the Tofts model, the \ac{aif} concentration $C_p(t)$ can be computed either from the patient or a population.
        In the two previous sections, we presented the algorithms that
        allow for the estimation of the patient-based \ac{aif} concentration.
        For a comparison with the previous approach, we also computed a population-based \ac{aif} that will be used later to compare the performance of both approaches.
        In this regard, the population-based \ac{aif} was estimated in
        accordance with the method of~\citep{meng2010comparison} by
        fitting the average patient-based \ac{aif}s to the model
        of~\cite{parker2006experimentally} which is formulated as
        shown in Eq.\,\eqref{eq:parker}, as follows:
        \begin{equation}
          C_p(t) = \sum_{n=1}^{2} \frac{A_n}{\sigma_n \sqrt{2 \pi}} \exp\left(\frac{- (t- T_n)^2}{2\sigma_{n}^{2}}\right) + \frac{\alpha \exp(-\beta t)}{1 + \exp{-s (t - \tau)}} ,
          \label{eq:parker}
        \end{equation}
        \noindent where $A_n$, $T_n$, and $\sigma_n$ are the scaling
        constants, centers, and widths of the n\textsuperscript{th}
        Gaussian, respectively; $\alpha$ and $\beta$ are the amplitude
        and decay constant of the exponential, respectively; and $s$ and $\tau$ are the width and center of the sigmoid function, respectively.
\end{description}

The parameters are estimated by fitting the model using constrained non-linear least-squares optimization, solved with the Trust Region Reflective algorithm~\citep{sorensen1982newton} and bounding the parameters to be positive.

@@

\citeauthor{gliozzi2011phenomenological} showed that the \ac{pun} approach can be used for \ac{dce}-\ac{mri} analysis~\citep{gliozzi2011phenomenological}.
The model has been successfully used in a \ac{cad} system proposed by~\cite{giannini2015fully}.
This model can be expressed as in Eq.\,\eqref{eq:pun}:

@@

\begin{equation}
  s(t) = \begin{cases}
    S_0 & 0 \leq t \leq t_0 \\
    S_M - (S_M - S_0) \exp\left( \frac{-(t - t_0)}{\tau} \right) & t_0 < t \leq t_0 + 2 \tau \\
    S_M - (S_M - S_0) \exp\left( \frac{-(t - t_0)}{\tau} \right) + w(t - t_0 + 2 \tau) & t > t_0 + 2 \tau
  \end{cases}
  \label{eq:huisman}
\end{equation}

@@

Instead of these successive estimations, we propose a unified optimization in which $t_0$ is fixed since it is a key parameter.
Therefore, $t_0$ is robustly estimated from the \ac{aif} signal since this is the most enhanced signal in which the start of enhancement is easily identifiable.
The \ac{aif} signal is computed as discussed in Section~\ref{sec:tofts}.
$t_0$ is estimated by finding the maximum of the first derivative of the \ac{aif} signal, always occurring at the beginning of the signal.
Then, the function in Eq.\,\eqref{eq:huisman} is fitted using non-linear least squares with the Trust Region Reflective algorithm~\citep{sorensen1982newton}.
Furthermore, the parameters $\tau$ and $S_M$ are bounded during the optimization to ensure robust estimations.
$\tau$ is bounded between $t_0$ and $t_f$, which is the time of the
last sample, and $S_M$ is bounded between $S_0$ and $\max(s(t))$.

>>>>

In this work, we propose a method to normalize \ac{dce}-\ac{mri}
prostate data to reduce inter-patient variations, although this method can be applied to any \ac{dce}-\ac{mri} sequences.
In \ac{t2w}-\ac{mri}, these variations are characterized by a shift and a scaling of the intensities as illustrated by the intensity \ac{pdf} in Fig.\,\ref{fig:t2w}.
Therefore, these variations can be corrected using a $z$-score approach,--- i.e., normalizing the data by subtracting the mean and dividing by the standard deviation ---assuming that the data follow a specific distribution~\cite{lemaitre2016normalization}.

\begin{figure*}
  \centering
  \hspace*{\fill}
  \subfloat[]{\label{subfig:pathhist}\includegraphics[width=1\textwidth]{02_methods/figures/heatmaprep.pdf}} \hfill
  \hspace*{\fill}
  \\
  \hspace*{\fill}
  \subfloat[$I_0$: 117; $t_0$: 6\textsuperscript{th} serie; wider st. dev.]{\label{subfig:pat1}\includegraphics[width=.49\textwidth]{02_methods/figures/pat1_annotated.pdf}} \hfill
  \subfloat[$I_0$: 103; $t_0$: 4\textsuperscript{th} serie; narrower std. dev.]{\label{subfig:pat2}\includegraphics[width=.49\textwidth]{02_methods/figures/pat2_annotated.pdf}} \hfill
  \hspace*{\fill}
  \caption{\protect\subref{subfig:pathhist} Illustration of the heatmap
    representation: all \ac*{pdf}s of the prostate gland are
    concatenated together to build an heatmap;
    \protect\subref{subfig:pat1}-\protect\subref{subfig:pat2} Heatmap of 2 patients
    revealing  the three types of inter-patient variations: intensity shift
    ($\Delta_i$), time shift ($\Delta_t$), and intensity scale
    ($\alpha_i$).}
  \label{fig:heatmap}
\end{figure*}

@@

\begin{equation}\small
  w_{ij} = \begin{cases}
    \alpha \exp(1 - \frac{H(i)}{\max(H)})       & \text{if } x_j = x_i + 1 \text{ and } y_j = y_i, \\
    (1 - \alpha) \exp(1 - \frac{H(i)}{\max(H)}) & \text{if } x_j = x_i \text{ and } y_j = y_i + 1, \\
    0                                           & \text{otherwise},
  \end{cases}
  \label{eq:weight}
\end{equation}

@@

\begin{figure*}
  \centering
  \hspace*{\fill}
  \subfloat[\acs*{rmse} computed for each patient of our dataset.]{\label{fig:rmse}\includegraphics[width=.44\textwidth]{02_methods/figures/rmse.pdf}} \hfill
  \subfloat[\acs*{rmse} after alignment using the curve parametric model.]{\label{fig:rmseal}\includegraphics[width=.49\textwidth]{02_methods/figures/rmse_aligned.pdf}}
  \hspace*{\fill}
  \caption{Illustration of the correction of the time offset and the data dispersion.}
  \label{fig:curveal}
\end{figure*}

@@

In this section, we summarize the different methods that have been
used for the quantification of \ac{dce}-\ac{mri} for \ac{cap}
detection~\cite{lemaitre2015computer} and will be used for comparison in this work.
Furthermore, we would like to emphasize the following additional contributions for this section: (i) a novel automatic \ac{aif} estimation algorithm based on clustering and (ii) a simplified semi-quantitative method using constrained optimization.

@@

In the Brix model~\cite{brix1991pharmacokinetic}, the \ac{mri} signal intensity is assumed to be proportional to the media concentration.
Therefore, the model is expressed as shown in Eq.\,\eqref{eq:brix}, as
follows:

\begin{dmath}
  s_n(t) = 1 + A \left[ \frac{\exp(k_{el} t') - 1}{k_{ep}(k_{ep} - k_{el})} \exp(- k_{el} t) - \frac{\exp(k_{ep} t') - 1}{k_{el}(k_{ep} - k_{el})} \exp(- k_{ep} t) \right],
  \label{eq:brix}
\end{dmath}

@@

\begin{dmath}
  s_n(t) = 1 + \frac{A}{\tau} \left[ \frac{k_{ep} \left( \exp(k_{el} t') - 1 \right)}{k_{el}(k_{ep} - k_{el})} \exp(- k_{el} t) - \frac{\exp(k_{ep} t') - 1}{(k_{ep} - k_{el})} \exp(- k_{ep} t) \right] ,
  \label{eq:hoffmann}
\end{dmath}

@@

\begin{figure*}
  \centering
  \hspace*{\fill}
  \subfloat[Original image.]{\label{fig:org}\includegraphics[width=.3\textwidth]{02_methods/figures/original.pdf}} \hfill
  \subfloat[Candidates region after clustering.]{\label{fig:cand}\includegraphics[width=.3\textwidth]{02_methods/figures/candidate.pdf}} \hfill
  \subfloat[Regions selected after applying the different criteria.]{\label{fig:final}\includegraphics[width=.3\textwidth]{02_methods/figures/aif.pdf}}
  \hspace*{\fill}
  \caption{Illustration of the segmentation of the area used to determine the \acs*{aif}.}
  \label{fig:aif}
\end{figure*}

\begin{description}
  \item[Segmentation of artery voxels and \ac{aif} estimation] The \ac{aif} signal from \ac{dce}-\ac{mri} can be manually estimated by selecting the most-enhanced voxels from the femoral or iliac arteries~\cite{meng2010comparison}.
    Few methods have been proposed to address the automated extraction
    of the \ac{aif} signal.
    \citeauthor{chen2008automatic} successively filtered the possible
    candidates to be considered as the \ac{aif} such that~\cite{chen2008automatic}:
    (i) dynamic signals with a small peak and voxels with a small wash-in are rejected by thresholding,
    (ii) a blob detector is used and large enough regions are maintained, and
    (iii) circular and cylindricality criteria are used to reject the false positives.
    \citeauthor{zhu2011automated} proposed an iterative method that
    involves the selection of voxels that best fit a gamma variate function~\cite{zhu2011automated}.
    However, this method requires the computation of the first and
    second derivatives as well as the maximum curvature points.
    \citeauthor{shanbhag2012generalized} proposed the following 4-step algorithm~\cite{shanbhag2012generalized,fennessy2015quantitative}:
    (i) remove the slices with artifacts and find the best slices based on intrinsic anatomic landmarks and enhancement characteristics,
    (ii) find the voxel candidates using the maximum enhanced voxels and a multi-label maximum entropy based thresholding algorithm,
    (iii) exclude the region next to the endorectal coil, and
    (iv) select the best 5 candidates that meet the enhancement characteristics and are correlated.

    All the above methods are rather complex, therefore,  we propose a simpler method that is based on the following reasonable assumptions:
    (i) all possible \ac{aif} signal candidates should have a similar shape,
    (ii) they should all have high enhancement, and
    (iii) the arteries should be almost round and within a size range.
    Therefore, each slice is clustered into regions using K-means clustering with $k=6$.
    The cluster made of the most enhanced signals is selected since it contains the artery signals.
    In this regards, the selection criteria corresponds to the 90\textsuperscript{th} percentile of the maximum \ac{dce}-\ac{mri} signal.
    Finally, regions with an eccentricity smaller than $0.5$ and an area in the range of $[100, 400]$ voxels are kept.
    Additionally, to remove voxels contaminated by the partial volume effect, only the $10\%$ most enhanced voxels of the possible candidates are kept as proposed by~\cite{schabel2008uncertainty} and the average signal is computed.
    A summary of the different segmentation steps is presented in Fig.\,\ref{fig:aif}.
    \item[Conversion of \ac{mri} signal intensity to concentration] To estimate the free parameters of the Tofts model (see Eq.\,\eqref{eq:exttofts}), the concentrations $C_t(t)$ and $C_p(t)$ need to be computed from the \ac{mri} signal intensity and the \ac{aif} signal, respectively.
      This conversion is based on the equation of the FLASH sequence\textemdash see~\ref{app:signaltoconc} for details\textemdash and is formulated as in Eq.\,\eqref{eq:conv}:
      \begin{equation}
        c(t) = \frac{1}{TR \cdot r_1} \ln\left( \frac{1 - \cos \alpha \cdot S^{*}\frac{s(t)}{S_0}}{1 - S^{*}\frac{s(t)}{S_0}} \right) - \frac{R_{10}}{r_1} ,
        \label{eq:conv}
      \end{equation}
      \noindent with,
      \begin{equation}
        S^{*} = \frac{1 - \exp(- TR \cdot R_{10})}{1 - \cos \alpha \cdot \exp(- TR \cdot R_{10})} ,
        \label{eq:sstarconv}
      \end{equation}
      \noindent where $s(t)$ is the \ac{mri} signal, $S_0$ is the \ac{mri} signal prior to the injection of the contrast media, $\alpha$ is the flip angle, $TR$ is the \acf{tr}, $R_{10}$ is the pre-contrast tissue relaxation time also equal to $\frac{1}{T_{10}}$, and $r_1$ is the relaxivity coefficient of the contrast agent.

      $T_{10}$ can be estimated from the acquisition of a T$_1$ map.
      However, this modality is not part of the clinical trial in this research and the value of $T_{10}$ is fixed to \SI{1600}{\ms} for both blood and prostate, in accordance with the values found in the literature~\cite{fennessy2015quantitative,de2004mr,carr2011magnetic}.
      \item[Estimation of population-based \ac{aif}] While estimating
        the pharmacokinetic parameters using the Tofts model, the \ac{aif} concentration $C_p(t)$ can be computed either from the patient or a population.
        In the two previous sections, we presented the algorithms that
        allow for the estimation of the patient-based \ac{aif} concentration.
        For a comparison with the previous approach, we also computed a population-based \ac{aif} that will be used later to compare the performance of both approaches.
        In this regard, the population-based \ac{aif} was estimated in
        accordance with the method of~\cite{meng2010comparison} by
        fitting the average patient-based \ac{aif}s to the model
        of~\cite{parker2006experimentally} which is formulated as
        shown in Eq.\,\eqref{eq:parker}, as follows:
        \begin{equation}\footnotesize
          C_p(t) = \sum_{n=1}^{2} \frac{A_n}{\sigma_n \sqrt{2 \pi}} \exp\left(\frac{- (t- T_n)^2}{2\sigma_{n}^{2}}\right) + \frac{\alpha \exp(-\beta t)}{1 + \exp{-s (t - \tau)}} ,
          \label{eq:parker}
        \end{equation}
        \noindent where $A_n$, $T_n$, and $\sigma_n$ are the scaling
        constants, centers, and widths of the n\textsuperscript{th}
        Gaussian, respectively; $\alpha$ and $\beta$ are the amplitude
        and decay constant of the exponential, respectively; and $s$ and $\tau$ are the width and center of the sigmoid function, respectively.
\end{description}

The parameters are estimated by fitting the model using constrained non-linear least-squares optimization, solved with the Trust Region Reflective algorithm~\cite{sorensen1982newton} and bounding the parameters to be positive.

@@

\citeauthor{gliozzi2011phenomenological} showed that the \ac{pun} approach can be used for \ac{dce}-\ac{mri} analysis~\cite{gliozzi2011phenomenological}.
The model has been successfully used in a \ac{cad} system proposed by~\cite{giannini2015fully}.
This model can be expressed as in Eq.\,\eqref{eq:pun}:

@@

\begin{equation}\tiny
  s(t) = \begin{cases}
    S_0 & 0 \leq t \leq t_0 \\
    S_M - (S_M - S_0) \exp\left( \frac{-(t - t_0)}{\tau} \right) & t_0 < t \leq t_0 + 2 \tau \\
    S_M - (S_M - S_0) \exp\left( \frac{-(t - t_0)}{\tau} \right) + w(t - t_0 + 2 \tau) & t > t_0 + 2 \tau
  \end{cases}
  \label{eq:huisman}
\end{equation}

@@

Instead of these successive estimations, we propose a unified optimization in which $t_0$ is fixed since it is a key parameter.
Therefore, $t_0$ is robustly estimated from the \ac{aif} signal since this is the most enhanced signal in which the start of enhancement is easily identifiable.
The \ac{aif} signal is computed as discussed in Section~\ref{sec:tofts}.
$t_0$ is estimated by finding the maximum of the first derivative of the \ac{aif} signal, always occurring at the beginning of the signal.
Then, the function in Eq.\,\eqref{eq:huisman} is fitted using non-linear least squares with the Trust Region Reflective algorithm~\cite{sorensen1982newton}.
Furthermore, the parameters $\tau$ and $S_M$ are bounded during the optimization to ensure robust estimations.
$\tau$ is bounded between $t_0$ and $t_f$, which is the time of the
last sample, and $S_M$ is bounded between $S_0$ and $\max(s(t))$.
